Predicting the spread and persistence of genetically modified dominant sterile male mosquitoes

### 2021-01-28 inst/manuscript/Draft/bmc_article_v9.tex
Response to FNIH/TM comments
Have been through FNIH and TM comments

In this analysis, we assume that village compounds (houses) provide a source of CO$_2$ that acts as an attractant for female \emph{Anopheles gambiae sensu strictu} and \emph{Anopheles coluzzi} mosquitoes \cite{DeJong1995, Majeed2017}. We further assume that male mosquitoes swarm in places where the probability of encountering receptive females is highest \citep{Diabate2015}, hence swarms occur in and around compounds (as evident in Figure 1 of \citep[][]{Epopa2017}). Consequently, high concentrations of attractant result in flights that (on average) are closer to the direction of the nearest compound for both males and females. Conversely, low concentrations have less influence on the mosquitoes and result in greater variability in flight direction. We do not incorporate the effect of wind speed or direction into the analysis (we assume the local dispersal dynamics are not greatly affected by the wind at this spatial scale), and also assume that the attraction process is not influenced by the number of mosquitoes.

@@

For this analysis we carefully elicited opinions on the mortality rate and dispersal distance of WT and DSM male mosquitoes from four independent experts who were not involved in the development of the DSM mosquitoes \cite{Hayes2015a}. Prior to the elicitation, experts were provided information on the genetic construct (originally incorporated into the G3 laboratory colony strain of \textit{An. gambiae}) and were told that the genetically modified strain would be repeatedly backcrossed with wild type mosquitoes captured in the vicinity of a biosecure laboratory in order to gradually replace the G3 genetic background with that of the locally originated strain. The location of the laboratory was provided to each expert but the species of local WT mosquitoes was not specified. For the mortality rate, experts were proposed to answer questions about either the probability of mortality, the probability of survival, or the life expectancy. All responses were converted to an estimate of the daily probability of mortality which was then converted to the mortality rate parameter used in the PDE model.

@@

\subsection{Spread and persistence of male DSM mosquitoes}
The predicted persistence of DSM mosquitoes following a release of a single cohort of 5000 males is summarised in Figure \ref{fig:plotTimeMoxMaxLoc}. The large uncertainty captured in Table \ref{tble:sumStatPPpostbma} is clearly reflected in these predictions: the mean expected survival is estimated to be about 9 to 10 days, with a 90\% probability that the expected number of male DSM mosquitoes falls below 1 by Day 10. The 90\% credible interval, however, is large, and there exists a very low probability ($\sim 0.05$) that it could take as long about two months for this to occur.

@@

\section{Discussion}
In this analysis we have used a combination of mathematical modeling, Bayesian inference, and expert elicitation to predict the spread and persistence of genetically modified DSM mosquitoes following the release of a single cohort of $5000$ individuals. These predictions were made as part of an independent risk assessment that was completed in May 2018 prior to the proposed field release. The field release itself occurred in July 2019 (\url{https://targetmalaria.org/target-malaria-proceeded-with-a-small-scale-release-of-genetically-modified-sterile-male-mosquitoes-in-bana-a-village-in-burkina-faso/}). We anticipate that the results from this field experiment will soon be forthcoming and provide an important opportunity to test the accuracy of the predictions presented here.

@@

Our posterior estimates of WT dispersal are conditional on a series of MRR experiments that were conducted prior to the elicitation of the dispersal prior distributions. These priors played no part in the experimental design, and the difference between the prior and posterior distributions depends on this design. This difference may reflect conservative prior estimates of mosquitoes' mean daily dispersal when in the vicinity of attractants (swarm locations) but it may also be an artefact of the MRR design that focussed all recaptures in and around village compounds. As \cite{Epopa2017} notes further intensive sampling studies outside or around villages will be useful to ensure that posterior estimates of dispersal are not blind to long range dispersal outcomes that are not witnessed because of the recapture site design.

>>>>

In this analysis, we assume that village compounds (groups of closely spaced houses) provide a source of CO$_2$ that acts as an attractant for female \emph{Anopheles gambiae sensu strictu} and \emph{Anopheles coluzzi} mosquitoes \cite{DeJong1995, Majeed2017}. We further assume that male mosquitoes swarm in places where the probability of encountering receptive females is highest \citep{Diabate2015}, hence swarms occur in and around compounds (as evident in Figure 1 of \citep[][]{Epopa2017}). Consequently, high concentrations of attractant result in flights that (on average) are closer to the direction of the nearest compound for both males and females. Conversely, low concentrations have less influence on the mosquitoes and result in greater variability in flight direction. We do not incorporate the effect of wind speed or direction into the analysis (we assume the local dispersal dynamics are not greatly affected by the wind at this spatial scale), and also assume that the attraction process is not influenced by the number of mosquitoes.

@@

For this analysis we carefully elicited opinions on the mortality rate and dispersal distance of WT and DSM male mosquitoes from four independent experts who were not involved in the development of the DSM mosquitoes \cite{Hayes2015a}. Prior to the elicitation, experts were provided information on the genetic construct (originally incorporated into the G3 laboratory colony strain of \textit{An. gambiae}) and were told that the genetically modified strain would be repeatedly backcrossed with wild type mosquitoes captured in the vicinity of a biosecure laboratory in order to gradually replace the G3 genetic background with that of the locally originated strain. The location of the laboratory was provided to each expert but the species of local WT mosquitoes was not specified. For the mortality rate, experts were proposed to answer questions about either the probability of mortality, the probability of survival, or the life expectancy. All responses were converted to an estimate of the daily probability of survival ($p$) which was then converted to the mortality rate parameter ($\mu = -\log(p)$) used in the PDE model.

@@

\subsection{Spread and persistence of male DSM mosquitoes}
The predicted persistence of DSM mosquitoes following a release of a single cohort of 5000 males is summarised in Figure \ref{fig:plotTimeMoxMaxLoc}. The large uncertainty captured in Table \ref{tble:sumStatPPpostbma} is clearly reflected in these predictions: the mean expected survival is estimated to be about 9 to 10 days, with a 90\% probability that the expected number of male DSM mosquitoes falls below 1 by Day 10. The 90\% credible interval, however, is large, and there is a small probability ($\sim 0.05$) that it could take as long about two months for this to occur.

@@

\section{Discussion}
In this analysis we have used a combination of mathematical modeling, Bayesian inference, and expert elicitation to predict the spread and persistence of genetically modified DSM mosquitoes following the release of a single cohort of $5000$ individuals. These predictions were made as part of an independent risk assessment that was completed in May 2018 prior to the proposed field release \citep{Hayes2015a, Hayes2018}. The field release itself occurred in July 2019 (\url{https://targetmalaria.org/target-malaria-proceeded-with-a-small-scale-release-of-genetically-modified-sterile-male-mosquitoes-in-bana-a-village-in-burkina-faso/}). We anticipate that the results from this field experiment will soon be forthcoming and provide an important opportunity to test the accuracy of the predictions presented here.

@@

Our posterior estimates of WT dispersal are conditional on a series of MRR experiments that were conducted prior to the elicitation of the dispersal prior distributions. These priors played no part in the experimental design, and the difference between the prior and posterior distributions depends on this design. This difference may reflect conservative prior estimates of mosquitoes' mean daily dispersal when in the vicinity of attractants (swarm locations) but it may also be an artefact of the MRR design that focussed efforts in and around village compounds to maximise the number of recaptures. As \cite{Epopa2017} notes, however, further intensive sampling studies outside or around villages will be useful to ensure that posterior estimates of dispersal are not blind to long range dispersal outcomes that are not witnessed because of the recapture site design.
